Double Standards in Judging Collective Action\thanks{}

### 2023-05-02 Documents/manuscript.qmd
first draft

Partisan differences can result from two social psychological processes. On the one hand, political partisanship reflects ideological differences in resistance to social change and acceptance of inequality which, in turn, are motivated by epistemic, existential, and relational needs [@jost_political_2003; @jost_left_2021]. As motivated cognition, ideological orientations shape how we think and feel about issues, actions, and events. On the other hand, political partisanship functions as a social identity [@finkel_political_2020; @osborne_psychology_2022] with evidence suggesting that partisans' policy preferences depends more on the stated position of their party than policy content [@cohen_party_2003]. Like other identities, political identities result in group-serving biases in how we evaluate ingroup and outgroup members. In practice, the two processes are often difficult to disentangle.

@@

Our research was motivated by real-world examples that seemed to suggest that where people draw the line between acceptable and unacceptable protest actions depends on *who* the protesters are and *what* they are protesting---in other words, that there are double standards in judging collective action. Our findings, however, showed that it is not so simple. Contradicting our hypothesis of identity-based double standards, participants in three preregistered experiments did not show consistent ingroup bias---in terms of class, race, and gender---when judging controversial protest actions taken by ingroup and outgroup members to advance their group's interests. Instead, supporting our hypothesis of ideology-based double standards, participants judged the same controversial protest actions to be more acceptable when the protesters' cause---protecting workers' rights, defunding the police, restricting abortion---aligned with their own ideological position. <!-- This suggests that where observers draw the line between acceptable and unacceptable forms of collective action depends not so much on the protesters' and the observers' identities, but on the protesters' cause and how it aligns with the observers' political and ideological orientation. --> In the remainder of this General Discussion, we consider some limiting conditions on our findings and discuss theoretical and practical implications for understanding the often divided response to collective action.

@@

Growing evidence shows that, in many situations, observers apply a more lenient standards when judging moral transgressions by ingroup members [@abrams_double_2013; @endevelt_everyone_2021]. And yet, our findings suggest that, unlike judgments in other domains, judgments about collective action are not subject to double standards based on group identities such as class, race, or gender.

One reason for this discrepancy might be that participants did not identify strongly enough with the social categories examined in our studies. This explanation, however, is not plausible because, first, we examined social categories (class, race, gender) that are central to most people's identities and because, second, we did not find evidence for ingroup bias even among highly identified group members. Another reason for this discrepancy might be that participants did not perceive the protesters' causes as relevant to the social identities examined in our studies. This explanation, however, conflicts with participants' reactions to the manipulation which, at least in Experiment 1--2, showed that participants understood the vignette in ingroup--outgroup terms.

<!-- Instead, I should offer three explanations for the observed partisan biases: (1) ideology (Jost, 2020, 2021), (2) identity, and (3) moralizaton. I could close by suggesting experiments to tease them apart. Assymmetry is evidence against identity-based processes as we would expect ingroup bias in both directions. -->

A more plausible explanation might be that collective action, even in its controversial forms, is not judged as a moral transgression [as in previous research on identity-based double standards, e.g., @abrams_double_2013]. Instead, observers might understand collective action primarily as a means to an end and judge how acceptable a form of collective action is in relation to its end ("the end justifies the mean"). This aligns with our finding that how acceptable participants judged protest actions to be depended on the protesters' cause, not their group membership.

Our findings showed that progressives and, to a lesser extent, conservatives consider the same controversial protest actions to be more acceptable when the protesters' cause aligns with their ideological position. So far, we explained this ideology-based double standard as resulting from conservatives being motivated to defend the system, progressives being motivated to challenge the system, and both being motivated to support collective action to those effects [@jost_missing_2017; @osborne_protesting_2019]. Other explanations, however, are plausible.

First, we might instead understand those double standards as an expression of ingroup bias based on an opinion-based [@mcgarty_collective_2009] or partisan [@finkel_political_2020] identity. Opinion-based identities are often a stronger predictor of collective action than identities based on other social categories [@bliuc_opinion-based_2007]. This aligns with Verkuyten et al.'s [-@verkuyten_intolerance_2022] finding that participants were more tolerant of the same transgressive protest actions when taken by their most-liked, rather than their least-liked, political group (e.g., climate activists).

Second, we might argue that their different moral concerns [@graham_liberals_2009; @kivikangas_moral_2021] lead progressives and conservatives to see their support of certain causes as a fundamental matter of right or wrong [@skitka_psychology_2021] and, by thus moralizing those causes, to accept more extreme means to achieving them. This aligns with Richardson & Conway's [-@richardson_standing_2022] finding that their different moral concerns explained why liberals rated protesting for liberal causes as more moral than conservatives and vice versa. Future research should seek to disentangle which of those plausible mechanisms best explains partisan differences in judging collective action.

Our findings raised the question why double standards in judging collective action were more pronounced among liberal than among conservative participants. One reason for this ideological asymmetry might be that collective action that violates prevailing norms or laws is seen as inherently threatening to the status quo, even if its aim is to bolster and defend the system. Political conservatism can be understood as having two psychologically related but distinguishable dimensions, resistance to change and justification of inequality [@jost_political_2003]. In this way, conservatives might be expected to condemn disruptive protest even if the protest's cause is to defend inequality. Future research should seek to establish how consistent this ideological asymmetry is across different political causes and cultural contexts.

<!-- Our findings provided a relevant comparison to findings from a recent experiment in which White Americans viewed pictures of either peaceful or destructive Black Lives Matter protests and rated how legitimate (i.e., fair, reasonable, legitimate, and justified) they considered the pictured protests to be [@teixeira_white_2022]. In this experiment, more conservative participants rated destructive and, to a lesser extent, peaceful protests to be less legitimate than more liberal participants. Likewise, participants who identified more strongly with their ethnic ingroup rated destructive and, to a lesser extent, peaceful protests to be less legitimate than participants who identified less strongly with their ingroup. However, participants' group identification was no longer associated with their protest ratings after controlling for their political orientation.^[These findings are based on our reanalysis, reported in the SOM, of Teixeira et al.'s [-@teixeira_white_2022] experiment. We thank the authors for making their data available online.] In this way, results from the experiment aligned with our conclusion that political ideologies are more important than social identities for shaping how people judge controversial forms of collective action. -->

Most broadly, our findings showed that, for many controversial protest actions, there is no consensus on what is and is not an acceptable means to advance a cause. Instead, this judgment seems to depend on what the cause of a protest is and how it aligns with one's beliefs and values. In this way, our results confirm our contention that the distinction between normative and non-normative forms of collective action should be situated in the eye of the beholder. By developing an instrument for measuring where people draw the line between acceptable and unacceptable forms of collective action, we provide a paradigm that we hope will stimulate research on when and why people differ in making this distinction.

After all, this distinction is often consequential. If people believe that even extreme actions are acceptable means to advance a cause they support, especially if they moralize their support for that cause [@mooijman_moralization_2018], they might support violent extremism. In this way, our findings could explain why 45% of Republicans supported the storming of the U.S. Capitol on January 6, 2021 [@yougov_capitol_2021] even though it was illegal, destructive, and violent. If people believe that even non-violent actions are unacceptable means to advance a cause they oppose, they might support stifling dissent. In this way, our findings could explain why Republicans responded to Black Lives Matter protests by seeking to further criminalize disruptive protest [e.g., blocking highways, @pew_republicans_2021] that might be most effective at challenging injustice [@shuman_disrupting_2021; @shuman_protest_2022]. For those reasons, we  encourage researchers to use our paradigm to further investigate when and why people apply different standards when judging collective action for different causes.

@@

Opinion polls show deep divides not only about the ends of protests, but also about the means of protest. For example, only 11\% of Republicans and 29\% of White Americans---compared to 59\% of Democrats and 66\% of Black Americans---considered kneeling during the national anthem---a non-violent form of collective action---an appropriate means to protest racist police violence [@yougov_nfl_2017]. Our research showed that, in contrast to well-documented double standards in other domains, people do not apply a different standard when judging collective action by members of the same class, race or gender or for a cause aligned with their own group's interests. Instead, our research demonstrated that people consider the same protest actions to be more acceptable when the cause of the protest aligns with their own ideological positions---in other words, that there are partisan double standards in judging collective action.

>>>>

Partisan differences can result from two social psychological processes. On the one hand, political partisanship reflects ideological differences in resistance to social change and acceptance of inequality which, in turn, are motivated by epistemic, existential, and relational needs [@jost_political_2003; @jost_left_2021]. As motivated cognition, ideological orientations shape how we think and feel about issues, actions, and events. On the other hand, political partisanship functions as a social identity [@finkel_political_2020; @osborne_psychology_2022] with evidence suggesting that partisans' policy preferences depend more on the stated position of their party than policy content [@cohen_party_2003]. Like other identities, political identities result in group-serving biases in how we evaluate ingroup and outgroup members. In practice, the two processes are often difficult to disentangle.

@@

Our research was motivated by real-world examples that seemed to suggest that where people draw the line between acceptable and unacceptable protest actions depends on *who* the protesters are and *what* they are protesting---in other words, that there are double standards in judging collective action. Our findings, however, showed that it is not so simple. Contradicting our hypothesis of identity-based double standards, participants in three preregistered experiments did not show consistent ingroup bias---in terms of class, race, and gender---when judging controversial protest actions taken by ingroup and outgroup members to advance their group's interests. Instead, supporting our hypothesis of ideology-based double standards, participants judged the same controversial protest actions to be more acceptable when the protesters' cause---protecting workers' rights, defunding the police, restricting abortion---aligned with their own ideological position. <!-- This suggests that where observers draw the line between acceptable and unacceptable forms of collective action depends not so much on the protesters' and the observers' identities, but on the protesters' cause and how it aligns with the observers' political and ideological orientation. --> In the remainder of this General Discussion, we consider limiting conditions on our findings and discuss theoretical and practical implications for understanding the often divided response to collective action.

@@

Growing evidence shows that, in many situations, observers apply a more lenient standards when judging moral transgressions by ingroup members [@abrams_double_2013; @endevelt_everyone_2021]. And yet, our findings suggested that, unlike judgments in other domains, judgments about collective action are not subject to double standards based on group identities such as class, race, or gender. We did not find evidence for those double standards even though we examined social categories that are central to most people's identities, even when we considered only highly identified group members, and even though participants' reactions to the manipulation showed that, at least in Experiments 1--2, they understood the protesters' causes in ingroup--outgroup terms.

Instead, our findings showed that progressives and, to a lesser extent, conservatives judge the same controversial protest actions to be more acceptable when the protesters' cause aligns with their own ideological position. We consider social, political, and moral psychological explanations for the observed ideology-based double standards.

First, we might understand ideology-based double standards as an expression of ingroup bias based on opinion-based [@mcgarty_collective_2009] or partisan [@osborne_psychology_2022; @finkel_political_2020] identities. This explanation aligns with Verkuyten et al.'s [-@verkuyten_intolerance_2022] finding that participants were more tolerant of transgressive protest actions when taken by their most-liked, rather than their least-liked, political group (e.g., climate activists). An identity-based account, however, implies that both conservatives and progressives would show ingroup bias and, therefore, cannot explain the observed asymmetry in the two groups' double standards.

Second, we might explain ideology-based double standards, as we have done so far, as resulting from conservatives being motivated to defend the system, progressives being motivated to challenge the system, and both being motivated to support collective action to those effects [@jost_missing_2017; @osborne_protesting_2019]. An ideology-based account could also  explain the asymmetry in how progressives and conservatives judged collective action for various causes. Political conservatism can be understood as uniting two motives, to resist social change and to maintain social inequality [@jost_political_2003; @jost_left_2021]. If conservatives perceive disruptive protest actions as inherently threatening to the social order, they will be motivated to condemn such actions even when the protesters' cause supports the unequal status quo (e.g., protesting *against* defunding the police).

Lastly, we might consider that collective action, even in its controversial forms, is not judged as a moral transgression [as in previous research on identity-based double standards, e.g., @abrams_double_2013]---but is instead understood as a means to an end and judged in relation to its end ("the end justifies the mean"). Their different moral concerns [@graham_liberals_2009; @kivikangas_moral_2021] might lead progressives and conservatives to see their support of certain causes as a fundamental matter of right or wrong [@skitka_psychology_2021] and, by thus moralizing those causes, to accept more extreme means to achieving them. This explanation aligns with Richardson & Conway's [-@richardson_standing_2022] finding that moral concerns explained why liberals rated protesting for liberal causes as more moral than conservatives (and vice versa). A values-based account could also explain why conservatives who, more so than liberals, endorse moral concerns related to loyalty and authority  tend to reject disruptive protest actions even when the protesters' cause otherwise aligns with their moral concerns.

Future research should, across different causes, contexts, and cultures, seek to disentangle which of the proposed social, political, and moral psychological processes best explains partisan differences in judging collective action.

Most broadly, our findings showed that, for many controversial protest actions, there is no consensus on whether they are acceptable means to advance a cause. <!--Instead, this judgment seems to depend on what the cause of a protest is and how it aligns with one's beliefs and values. -->This confirms our contention that the distinction between normative and non-normative forms of collective action should be situated in the eye of the beholder. By developing an instrument for measuring where people draw the line between acceptable and unacceptable forms of collective action, we provide a paradigm that we hope will stimulate research on when and why people differ in making this distinction.

After all, this distinction is consequential. If people believe that even extreme actions are acceptable means to advance a cause they support, especially if they moralize their support for that cause [@mooijman_moralization_2018], they might support violent extremism. In this way, our findings could explain why 45% of Republicans supported the storming of the U.S. Capitol on January 6, 2021 [@yougov_capitol_2021] even though it was illegal, destructive, and violent. If people believe that even non-violent actions are unacceptable means to advance a cause they oppose, they might support stifling dissent. In this way, our findings could explain why Republicans responded to Black Lives Matter protests by seeking to further criminalize disruptive protest [e.g., blocking highways, @pew_republicans_2021] that might be most effective at challenging injustice [@shuman_disrupting_2021; @shuman_protest_2022].<!-- For those reasons, we  encourage researchers to use our paradigm to further investigate when and why people apply different standards when judging collective action for different causes.-->

@@

Opinion polls show deep divides not only about the ends of protests, but also about the means of protest. For example, only 11\% of Republicans and 29\% of White Americans---compared to 59\% of Democrats and 66\% of Black Americans---considered kneeling during the national anthem---a non-violent form of collective action---an appropriate means to protest racist police violence [@yougov_nfl_2017]. Our research showed that, in contrast to well-documented double standards in other domains, people do not apply a different standard when judging collective action by members of the same class, race or gender or for a cause aligned with their own group's interests. Instead, our research demonstrated that people consider the same protest actions to be more acceptable when the cause of the protest aligns with their own ideological position<!-- ---in other words, that there are partisan double standards in judging collective action -->.

### 2023-05-02 Documents/manuscript.tex
first draft

Partisan differences can result from two social psychological processes.
On the one hand, political partisanship reflects ideological differences
in resistance to social change and acceptance of inequality which, in
turn, are motivated by epistemic, existential, and relational needs
(Jost et al., 2003; Jost, 2021). As motivated cognition, ideological
orientations shape how we think and feel about issues, actions, and
events. On the other hand, political partisanship functions as a social
identity (Finkel et al., 2020; Harris et al., 2022) with evidence
suggesting that partisans' policy preferences depends more on the stated
position of their party than policy content (Cohen, 2003). Like other
identities, political identities result in group-serving biases in how
we evaluate ingroup and outgroup members. In practice, the two processes
are often difficult to disentangle.

@@

Our research was motivated by real-world examples that seemed to suggest
that where people draw the line between acceptable and unacceptable
protest actions depends on \emph{who} the protesters are and \emph{what}
they are protesting---in other words, that there are double standards in
judging collective action. Our findings, however, showed that it is not
so simple. Contradicting our hypothesis of identity-based double
standards, participants in three preregistered experiments did not show
consistent ingroup bias---in terms of class, race, and gender---when
judging controversial protest actions taken by ingroup and outgroup
members to advance their group's interests. Instead, supporting our
hypothesis of ideology-based double standards, participants judged the
same controversial protest actions to be more acceptable when the
protesters' cause---protecting workers' rights, defunding the police,
restricting abortion---aligned with their own ideological position. In
the remainder of this General Discussion, we consider some limiting
conditions on our findings and discuss theoretical and practical
implications for understanding the often divided response to collective
action.

@@

Growing evidence shows that, in many situations, observers apply a more
lenient standards when judging moral transgressions by ingroup members
(Abrams et al., 2013; Endevelt et al., 2021). And yet, our findings
suggest that, unlike judgments in other domains, judgments about
collective action are not subject to double standards based on group
identities such as class, race, or gender.

One reason for this discrepancy might be that participants did not
identify strongly enough with the social categories examined in our
studies. This explanation, however, is not plausible because, first, we
examined social categories (class, race, gender) that are central to
most people's identities and because, second, we did not find evidence
for ingroup bias even among highly identified group members. Another
reason for this discrepancy might be that participants did not perceive
the protesters' causes as relevant to the social identities examined in
our studies. This explanation, however, conflicts with participants'
reactions to the manipulation which, at least in Experiment 1--2, showed
that participants understood the vignette in ingroup--outgroup terms.

A more plausible explanation might be that collective action, even in
its controversial forms, is not judged as a moral transgression (as in
previous research on identity-based double standards, e.g., Abrams et
al., 2013). Instead, observers might understand collective action
primarily as a means to an end and judge how acceptable a form of
collective action is in relation to its end (``the end justifies the
mean''). This aligns with our finding that how acceptable participants
judged protest actions to be depended on the protesters' cause, not
their group membership.

Our findings showed that progressives and, to a lesser extent,
conservatives consider the same controversial protest actions to be more
acceptable when the protesters' cause aligns with their ideological
position. So far, we explained this ideology-based double standard as
resulting from conservatives being motivated to defend the system,
progressives being motivated to challenge the system, and both being
motivated to support collective action to those effects (Jost et al.,
2017; Osborne et al., 2019). Other explanations, however, are plausible.

First, we might instead understand those double standards as an
expression of ingroup bias based on an opinion-based (McGarty et al.,
2009) or partisan (Finkel et al., 2020) identity. Opinion-based
identities are often a stronger predictor of collective action than
identities based on other social categories (Bliuc et al., 2007). This
aligns with Verkuyten et al.'s (2022) finding that participants were
more tolerant of the same transgressive protest actions when taken by
their most-liked, rather than their least-liked, political group (e.g.,
climate activists).

Second, we might argue that their different moral concerns (Graham et
al., 2009; Kivikangas et al., 2021) lead progressives and conservatives
to see their support of certain causes as a fundamental matter of right
or wrong (Skitka et al., 2021) and, by thus moralizing those causes, to
accept more extreme means to achieving them. This aligns with Richardson
\& Conway's (2022) finding that their different moral concerns explained
why liberals rated protesting for liberal causes as more moral than
conservatives and vice versa. Future research should seek to disentangle
which of those plausible mechanisms best explains partisan differences
in judging collective action.

Our findings raised the question why double standards in judging
collective action were more pronounced among liberal than among
conservative participants. One reason for this ideological asymmetry
might be that collective action that violates prevailing norms or laws
is seen as inherently threatening to the status quo, even if its aim is
to bolster and defend the system. Political conservatism can be
understood as having two psychologically related but distinguishable
dimensions, resistance to change and justification of inequality (Jost
et al., 2003). In this way, conservatives might be expected to condemn
disruptive protest even if the protest's cause is to defend inequality.
Future research should seek to establish how consistent this ideological
asymmetry is across different political causes and cultural contexts.

Most broadly, our findings showed that, for many controversial protest
actions, there is no consensus on what is and is not an acceptable means
to advance a cause. Instead, this judgment seems to depend on what the
cause of a protest is and how it aligns with one's beliefs and values.
In this way, our results confirm our contention that the distinction
between normative and non-normative forms of collective action should be
situated in the eye of the beholder. By developing an instrument for
measuring where people draw the line between acceptable and unacceptable
forms of collective action, we provide a paradigm that we hope will
stimulate research on when and why people differ in making this
distinction.

After all, this distinction is often consequential. If people believe
that even extreme actions are acceptable means to advance a cause they
support, especially if they moralize their support for that cause
(Mooijman et al., 2018), they might support violent extremism. In this
way, our findings could explain why 45\% of Republicans supported the
storming of the U.S. Capitol on January 6, 2021 (YouGov, 2021) even
though it was illegal, destructive, and violent. If people believe that
even non-violent actions are unacceptable means to advance a cause they
oppose, they might support stifling dissent. In this way, our findings
could explain why Republicans responded to Black Lives Matter protests
by seeking to further criminalize disruptive protest (e.g., blocking
highways, Quinton, 2021) that might be most effective at challenging
injustice (Shuman et al., 2022, 2021). For those reasons, we encourage
researchers to use our paradigm to further investigate when and why
people apply different standards when judging collective action for
different causes.

@@

Opinion polls show deep divides not only about the ends of protests, but
also about the means of protest. For example, only 11\% of Republicans
and 29\% of White Americans---compared to 59\% of Democrats and 66\% of
Black Americans---considered kneeling during the national anthem---a
non-violent form of collective action---an appropriate means to protest
racist police violence (YouGov, 2017). Our research showed that, in
contrast to well-documented double standards in other domains, people do
not apply a different standard when judging collective action by members
of the same class, race or gender or for a cause aligned with their own
group's interests. Instead, our research demonstrated that people
consider the same protest actions to be more acceptable when the cause
of the protest aligns with their own ideological positions---in other
words, that there are partisan double standards in judging collective
action.

@@

\leavevmode\vadjust pre{\hypertarget{ref-bliuc_opinion-based_2007}{}}%
Bliuc, A.-M., McGarty, C., Reynolds, K., \& Muntele, D. (2007).
Opinion-based group membership as a predictor of commitment to political
action. \emph{European Journal of Social Psychology}, \emph{37}(1),
19--32. \url{https://doi.org/10.1002/ejsp.334}

>>>>

Partisan differences can result from two social psychological processes.
On the one hand, political partisanship reflects ideological differences
in resistance to social change and acceptance of inequality which, in
turn, are motivated by epistemic, existential, and relational needs
(Jost et al., 2003; Jost, 2021). As motivated cognition, ideological
orientations shape how we think and feel about issues, actions, and
events. On the other hand, political partisanship functions as a social
identity (Finkel et al., 2020; Harris et al., 2022) with evidence
suggesting that partisans' policy preferences depend more on the stated
position of their party than policy content (Cohen, 2003). Like other
identities, political identities result in group-serving biases in how
we evaluate ingroup and outgroup members. In practice, the two processes
are often difficult to disentangle.

@@

Our research was motivated by real-world examples that seemed to suggest
that where people draw the line between acceptable and unacceptable
protest actions depends on \emph{who} the protesters are and \emph{what}
they are protesting---in other words, that there are double standards in
judging collective action. Our findings, however, showed that it is not
so simple. Contradicting our hypothesis of identity-based double
standards, participants in three preregistered experiments did not show
consistent ingroup bias---in terms of class, race, and gender---when
judging controversial protest actions taken by ingroup and outgroup
members to advance their group's interests. Instead, supporting our
hypothesis of ideology-based double standards, participants judged the
same controversial protest actions to be more acceptable when the
protesters' cause---protecting workers' rights, defunding the police,
restricting abortion---aligned with their own ideological position. In
the remainder of this General Discussion, we consider limiting
conditions on our findings and discuss theoretical and practical
implications for understanding the often divided response to collective
action.

@@

Growing evidence shows that, in many situations, observers apply a more
lenient standards when judging moral transgressions by ingroup members
(Abrams et al., 2013; Endevelt et al., 2021). And yet, our findings
suggested that, unlike judgments in other domains, judgments about
collective action are not subject to double standards based on group
identities such as class, race, or gender. We did not find evidence for
those double standards even though we examined social categories that
are central to most people's identities, even when we considered only
highly identified group members, and even though participants' reactions
to the manipulation showed that, at least in Experiments 1--2, they
understood the protesters' causes in ingroup--outgroup terms.

Instead, our findings showed that progressives and, to a lesser extent,
conservatives judge the same controversial protest actions to be more
acceptable when the protesters' cause aligns with their own ideological
position. We consider social, political, and moral psychological
explanations for the observed ideology-based double standards.

First, we might understand ideology-based double standards as an
expression of ingroup bias based on opinion-based (McGarty et al., 2009)
or partisan (Finkel et al., 2020; Harris et al., 2022) identities. This
explanation aligns with Verkuyten et al.'s (2022) finding that
participants were more tolerant of transgressive protest actions when
taken by their most-liked, rather than their least-liked, political
group (e.g., climate activists). An identity-based account, however,
implies that both conservatives and progressives would show ingroup bias
and, therefore, cannot explain the observed asymmetry in the two groups'
double standards.

Second, we might explain ideology-based double standards, as we have
done so far, as resulting from conservatives being motivated to defend
the system, progressives being motivated to challenge the system, and
both being motivated to support collective action to those effects (Jost
et al., 2017; Osborne et al., 2019). An ideology-based account could
also explain the asymmetry in how progressives and conservatives judged
collective action for various causes. Political conservatism can be
understood as uniting two motives, to resist social change and to
maintain social inequality (Jost et al., 2003; Jost, 2021). If
conservatives perceive disruptive protest actions as inherently
threatening to the social order, they will be motivated to condemn such
actions even when the protesters' cause supports the unequal status quo
(e.g., protesting \emph{against} defunding the police).

Lastly, we might consider that collective action, even in its
controversial forms, is not judged as a moral transgression (as in
previous research on identity-based double standards, e.g., Abrams et
al., 2013)---but is instead understood as a means to an end and judged
in relation to its end (``the end justifies the mean''). Their different
moral concerns (Graham et al., 2009; Kivikangas et al., 2021) might lead
progressives and conservatives to see their support of certain causes as
a fundamental matter of right or wrong (Skitka et al., 2021) and, by
thus moralizing those causes, to accept more extreme means to achieving
them. This explanation aligns with Richardson \& Conway's (2022) finding
that moral concerns explained why liberals rated protesting for liberal
causes as more moral than conservatives (and vice versa). A values-based
account could also explain why conservatives who, more so than liberals,
endorse moral concerns related to loyalty and authority tend to reject
disruptive protest actions even when the protesters' cause otherwise
aligns with their moral concerns.

Future research should, across different causes, contexts, and cultures,
seek to disentangle which of the proposed social, political, and moral
psychological processes best explains partisan differences in judging
collective action.

Most broadly, our findings showed that, for many controversial protest
actions, there is no consensus on whether they are acceptable means to
advance a cause. This confirms our contention that the distinction
between normative and non-normative forms of collective action should be
situated in the eye of the beholder. By developing an instrument for
measuring where people draw the line between acceptable and unacceptable
forms of collective action, we provide a paradigm that we hope will
stimulate research on when and why people differ in making this
distinction.

After all, this distinction is consequential. If people believe that
even extreme actions are acceptable means to advance a cause they
support, especially if they moralize their support for that cause
(Mooijman et al., 2018), they might support violent extremism. In this
way, our findings could explain why 45\% of Republicans supported the
storming of the U.S. Capitol on January 6, 2021 (YouGov, 2021) even
though it was illegal, destructive, and violent. If people believe that
even non-violent actions are unacceptable means to advance a cause they
oppose, they might support stifling dissent. In this way, our findings
could explain why Republicans responded to Black Lives Matter protests
by seeking to further criminalize disruptive protest (e.g., blocking
highways, Quinton, 2021) that might be most effective at challenging
injustice (Shuman et al., 2022, 2021).

@@

Opinion polls show deep divides not only about the ends of protests, but
also about the means of protest. For example, only 11\% of Republicans
and 29\% of White Americans---compared to 59\% of Democrats and 66\% of
Black Americans---considered kneeling during the national anthem---a
non-violent form of collective action---an appropriate means to protest
racist police violence (YouGov, 2017). Our research showed that, in
contrast to well-documented double standards in other domains, people do
not apply a different standard when judging collective action by members
of the same class, race or gender or for a cause aligned with their own
group's interests. Instead, our research demonstrated that people
consider the same protest actions to be more acceptable when the cause
of the protest aligns with their own ideological position.

@@
